\documentclass{article}
\usepackage{amsmath, amssymb,amsfonts,mdframed, tikz}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usetikzlibrary{shapes,arrows,calc,positioning,backgrounds}
\title{Algorithms \& Complexity: Lecture 8, Greedy Algorithms 2}

\author{Sam Barrett}

\newcommand{\T}{\texttt{T }}
\newcommand{\True}{\texttt{True }}
\newcommand{\F}{\texttt{F }}
\newcommand{\False}{\texttt{False }}
\newcommand{\NP}{\mathbf{NP}}
\renewcommand{\P}{\mathbf{P}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\R}[1]{\texttt{Req}(#1)}
\renewcommand{\S}[1]{\texttt{Start}(#1)}
\renewcommand{\F}[1]{\texttt{Finish}(#1)}

\newmdtheoremenv{lemma}{Lemma}
\newmdtheoremenv{proof}{Proof}
\newmdtheoremenv{definition}{Definition}
\newmdtheoremenv{theorem}{Theorem}
\newmdtheoremenv{problem}{Problem}
\newmdtheoremenv{example}{Example}

\begin{document}
\maketitle

\section{Interval Scheduling Problem}

\begin{problem}(The Interval Scheduling Problem)



  \begin{itemize}

    \item Given a set of $n$ requests $R = \left\{ \R{1}, \R{2}, \R{3}, \ldots, \R{n} \right\} $
  \item $\R{i}$ has a start time given by $\S{i}$ and a finish time by $\F{i}$
  \item There exists a machine which can handle one request at a time.
  \item Two requests \textit{conflict} if they overlap

  \end{itemize}

  Select a set $C\subseteq R$ of requests s.t. $|C|$ is maximised and no two requests from $C$ conflict.
\end{problem}


See Slides for lecture 8 for details on failed greedy algorithm construction.

A correct greedy algorithm for tackling this problem can be seen below:

\begin{algorithm}

  \caption{Greedy Interval Scheduling}


  Let $R = \{ \R{1},\R{2},\ldots,\R{i},\ldots,\R{n} \} $ be the set of all requests

  Let $C$ denote the set of requests that we select

  Instantiate $C = \emptyset $

  \While{$R \neq \emptyset$}{
    Find the request $\R{i} \in R$ which has the \textbf{smallest finish time}

    Add $\R{i}$ to $C$

    Delete from $R$ all requests that conflict with $\R{i}$
  }
\end{algorithm}

\subsection{Correctness}

\begin{itemize}
  \item $C$ does not contain any conflicting requests

        Suppose there is another set $\texttt{OPT}$ which selects more requests than $C$. Let $C$ select $\R{i_{1}}, \R{i_{2}}, \ldots,\R{i_{k}}$ in that order.

        Let $\texttt{OPT}$ schedule $\R{j_{1}},\R{j_{2}},\ldots, \R{j_{m}}$ in that order.

        \begin{lemma}
          For each $1 \leq l \leq k$, we have $\F{i_{l}} \leq \F{j_{l}}$
        \end{lemma}

        \begin{proof}

          \begin{enumerate}
            \item \textbf{Base Case:} $l=1$
            \item \textbf{Inductive Step:} $\S{j_{l}} \geq \F{j_{l-1}} \geq \F{i_{l-1}}$

                  Therefore, $\R{j_{l}}$ does \textbf{not} conflict with $\R{i_{l-1}}$. However, our algorithm chose $i_{l}$ instead, meaning $\F{i_{l}} \leq \F{j_{l}}$
          \end{enumerate}

        \end{proof}

        Since $m > k$, $\texttt{OPT}$ selects a request $\R{j_{k+1}}$ we can say:

        \[
        \S{j_{k+1}} \geq \F{j_{k}} \geq \F{i_{k}}
        \]

        And thus derive a contradiction as our algorithm stops after selecting $i_{k}$

\end{itemize}

A harder variant of the interval scheduling problem must schedule \textbf{all} requests, minimising lateness. And is defined as follows:

\begin{problem}

  \begin{itemize}
    \item We have $n$ requests
    \item Each requests has a duration given by $\texttt{Time}(i)$
    \item Additionally, each request, $\R{i}$ now had a deadline, $\texttt{Deadline}(i)$
    \item Choosing a start time $\S{i}$ for each request not gives a finish time $\F{i} = \S{i} + \texttt{Time}(i)$

    \item A request $\R{i}$ is \textbf{late} is $\F{i} > \texttt{Deadline}(i)$

          \[
          \texttt{Lateness}(i) =
          \begin{cases}
            \F{i} - \texttt{Deadline}(i) & \text{if } \F{i} > \texttt{Deadline}(i) \\
            0 & \text{otherwise}
          \end{cases}
          \]

  \end{itemize}

  Schedule all requests in a non-conflicting way, minimising the maximum lateness.
\end{problem}

A simple greedy algorithm does not find an optimal scheduling.

\end{document}
